---
title: "1_CReATe_data_cleaning"
author: "Katerina Placek"
date: "8/31/2020"
output: html_document
---
Clear workspace and set working directory.
```{r setup, include = TRUE, echo = FALSE}
rm(list = ls(all.names = TRUE)) 
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Katerina/Manuscripts/EMBO_MolMedicine/data_code_share/")
```

Load necessary packages for code.
```{r load_packages, echo = FALSE}
library(pacman)
p_load(tidyverse, nlme, circlize, ComplexHeatmap, tableone, readxl, reshape2, RColorBrewer, EnvStats, car, knitr, matrixStats, png, gridExtra, install = TRUE, update = getOption("pac_update"), character.only = FALSE)
```

Import CReATe data and clean data fields. Define cognitive impairment on the ECAS according to Abrahams et al. criteria.
```{r import_and_clean_data, warning = FALSE}
CReATe_data <- read.csv("data_notshared/CReATe_data.csv", stringsAsFactors = FALSE)[-1]

CReATe_data <- CReATe_data  %>% 
  mutate(ID = as.factor(ID), 
         VisitMonths = case_when(
            Visit == "Baseline" ~ 0,
            Visit == "3 months" ~ 3,
            Visit == "6 months" ~ 6,
            Visit == "12 months" ~ 12,
            Visit == "18 months" ~ 18,
            Visit == "2 years" ~ 24),
        Mutation_char = case_when(
            is.na(Mutation) ~ "None",
            Mutation == "C9ORF72 HET and UBQLN2 (P525S) HEM" ~ "C9ORF72/UBQLN2",
            grepl('SOD1', Mutation) ~ "SOD1", 
            grepl('TBK1', Mutation) ~ "TBK1", 
            grepl('TARDBP', Mutation) ~ "TARDBP", 
            grepl('C9ORF72', Mutation) ~ "C9ORF72",
            grepl('SQSTM1', Mutation) ~ "SQSTM1"),
        Mutation_bin = ifelse(is.na(Mutation), "No Mutation", "Mutation"),
        Educ_char = replace(Educ_char, Educ_char=="Unknown", "Other"),
        Educ_char = factor(Educ_char, levels = c("Grade School", "High School", "2-year College", "4-year College", "Master's", "Doctoral", "Other")), 
        Educ_bin = case_when(
          Educ_char=="2-year College"| Educ_char=="4-year College"| Educ_char== "Master's"| Educ_char== "Doctoral" ~ "College Education", 
          Educ_char== "Grade School"| Educ_char== "High School" | Educ_char=="Other" ~ "No College Education"),
        Male = as.character(Male)) %>%
  mutate(Lag = BslnAge - OnsetAge) %>% 
  mutate(C9ORF72 = ifelse(grepl('C9ORF72', Mutation_char), 1, 0),
         Other_Mutation = ifelse(Mutation_char == "None" | Mutation_char == "C9ORF72", 0, 1)) %>% 
  group_by(ID) %>% 
  mutate(nVisits = n()) %>% 
  ungroup()

CReATe_data <- CReATe_data %>%
  group_by(ID) %>%
  arrange(VisitNum) %>%
  slice(1L) %>%
  ungroup() %>%
  rename_at(vars(starts_with("ECAS_"), UMN_adj, LMN_adj,ALSFRS_TotalR_DM), funs(paste0("bl_",.))) %>% 
  mutate(ClinDx2 = ifelse(ClinDx == "ALS", case_when(
bl_ECAS_Verb <=14 | bl_ECAS_Exec <=33 | bl_ECAS_Lang <=26 ~ "ALSci",
    bl_ECAS_Verb >14 | bl_ECAS_Exec >33 | bl_ECAS_Lang >26 ~ "ALS"), ClinDx)) %>% 
  dplyr::select(ID, starts_with("bl_"), ClinDx2) %>% 
  right_join(., CReATe_data)
```

Make vectors of clinical and genetic variables. Remove 1 participant with a clinical diagnosis of FTD,  remove 7 participants with missing Lag information (time from symptom onset to baseline visit), and remove 1 participant with no available UMN score information.
```{r filter_missing_data}
clin_vars <- tidyselect::vars_select(names(CReATe_data), starts_with("ECAS_"), ALSFRS_TotalR_DM, UMN_adj, LMN_adj)

gen_vars <- tidyselect::vars_select(names(CReATe_data), starts_with("rs"), PC1, PC2, C9ORF72, Other_Mutation)

CReATe_data <- CReATe_data  %>% 
  filter(ClinDx != "FTD") %>%
  filter(!is.na(Lag)) %>%
  filter_at(gen_vars, all_vars(!is.na(.))) %>% 
  dplyr::select(ID, all_of(clin_vars)) %>% 
  group_by(ID) %>%
  summarise_all(list(~sum(!is.na(.)))) %>% 
  filter(UMN_adj>0) %>% 
  dplyr::select(ID) %>% 
  left_join(., CReATe_data)
```

Plot group counts of the total number of visits for participants in the CReATe PGB cohort.
```{r plot_nVisits, warning = FALSE}
CReATe_data %>% 
  dplyr::select(ID, VisitNum, nVisits) %>% 
  mutate(nVisits=as.factor(nVisits)) %>% 
  group_by(ID) %>%
  arrange(VisitNum) %>%
  slice(1L) %>% 
  ggplot(., aes(nVisits))+
  geom_histogram(stat = "count") +
  xlab("Number of Visits") +
  ylab("Number of Patients") +
  ggtitle("CReATe PGB Cohort: Longitudinal Timepoints") +
  theme_minimal()
```

Plot each participant's standardized performance at baseline on each clinical measure:
```{r plot_baseline, warning = FALSE}
CReATe_data %>% 
  group_by(ID) %>%
  arrange(VisitNum) %>%
  slice(1L) %>%
  ungroup() %>%
  dplyr::select(ID, one_of(clin_vars)) %>% 
  mutate_at(clin_vars, scale) %>% 
  gather(key = key, value = value, -ID) %>% 
  ggplot(., aes(key,value, colour = key))+
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(color=key), size=0.4, alpha=0.9) +
  theme_minimal()+
  theme(legend.position = "none", axis.text.x = element_text(angle = 90))+
  labs(y= "Standard deviation", x = "") +
  ggtitle("CReATe PGB Cohort - Clinical Performance at First Visit")
```

Plot each participant's standardized longitudinal performance on each clinical measure:
```{r plot_longitudinal, warning = FALSE}
CReATe_data %>% 
  filter(nVisits >1) %>% 
  dplyr::select(ID, VisitMonths, one_of(clin_vars)) %>% 
  gather(key = key, value = value, -ID, -VisitMonths) %>% 
  ggplot(., aes(VisitMonths, value, group = ID, colour = key))+
  geom_line(alpha = 0.5) +
  facet_wrap(~key, scales = "free_y")+
  theme_minimal()+
  ggtitle("CReATe PGB Cohort - Longitudinal Clinical Performance")+
  theme(legend.position = "none")
```

Remove 3 participants who performed > 5 standard deviations from the group mean on clinical variables at baseline and save out a CSV file for the cleaned longitudinal data
```{r filter_outliers, warning = FALSE}
CReATe_data <- CReATe_data %>% 
  group_by(ID) %>%
  arrange(VisitNum) %>%
  slice(1L) %>%
  ungroup() %>%
  dplyr::select(ID, one_of(clin_vars)) %>% 
  mutate_at(clin_vars, scale) %>% 
  gather(key = key, value = value, -ID) %>% 
  spread(key = key, value=value) %>%
  filter_at(clin_vars, all_vars(.>=-5 | is.na(.))) %>%
  dplyr::select(ID) %>% 
  left_join(., CReATe_data, by = "ID")

write.csv(CReATe_data, "data_notshared/cleaned_CReATe_data_long.csv")
```

Create a table of demographic information (Table 1) for the final CReATe PGB cohort. 
```{r Table1}
Table1 <- CReATe_data %>% 
  group_by(ID) %>%
  arrange(VisitNum) %>%
  slice(1L) %>%
  dplyr::select(ClinDx2,Male,OnsetAge,Mutation_char, Onset3_Bulbar,Onset3_Limb,Lag,Onset3_Other, Educ_bin, nVisits, starts_with("bl")) %>% 
  mutate(OnsetSite = case_when(
    Onset3_Bulbar==1 && Onset3_Other==0 && Onset3_Limb==0 ~ "Bulbar",
    Onset3_Bulbar==0 && Onset3_Other==0 && Onset3_Limb==1 ~ "Limb",
    Onset3_Bulbar==1 && Onset3_Other==0 && Onset3_Limb==1 ~ "Bulbar & Limb",
    Onset3_Bulbar==1 && Onset3_Other==1 && Onset3_Limb==0 ~ "Bulbar & Other",
    Onset3_Bulbar==0  && Onset3_Other==1 && Onset3_Limb==1 ~ "Limb & Other",
    Onset3_Bulbar==0  && Onset3_Other==1 && Onset3_Limb==0 ~ "Other"
  )) %>% 
  ungroup() %>% 
  mutate(OnsetSite = as.factor(OnsetSite)) %>% 
CreateTableOne(vars = colnames(.)[-1], data = ., strata = c("ClinDx2"))

print(Table1)
tab1export <- print(Table1, exact = "stage", quote = FALSE, noSpaces = TRUE, printToggle = FALSE)
write.csv(tab1export, file = "Table1.csv")
```

Run linear mixed effects models for each clinical variable including random effects for individual (intercept) and time*individual (slope). Extract individualized slopes and intercepts for each participant on each clinical variable, standardize values, and save to a data frame.
```{r lme_models, warning = FALSE}
models <- list()

for(v in 1:length(clin_vars)){
    formula <- as.formula(paste(clin_vars[v], "~ BslnAge + Educ_bin + Onset3_Bulbar + Lag + VisitMonths"))
    models[[v]] <- lme(fixed = formula, data=CReATe_data, random = reStruct( ~ VisitMonths | ID, REML=FALSE), na.action = na.exclude, control=lmeControl(opt='optim'))
}

mod <- matrix(nrow = length(ranef(models[[1]])$VisitMonths), ncol = 0)

names <- c()

for (v in 1:length(clin_vars)){
  mod <- cbind(mod, as.numeric(ranef(models[[v]])$VisitMonths), as.numeric(ranef(models[[v]])$`(Intercept)`))
  names <- c(names, paste0(clin_vars[[v]], "_slope"), paste0(clin_vars[[v]], "_intercept") )
}

mod <- as.data.frame(mod) %>% 
  mutate_all(scale) %>% 
  mutate_all(as.numeric)

names(mod) <- names

mod$ID <- row.names(coef(models[[1]]))

mod <- CReATe_data %>% 
  group_by(ID) %>%
  arrange(VisitNum) %>%
  slice(1L) %>%
  left_join(., mod, by = "ID")%>% 
  arrange(ClinDx2) %>% 
  ungroup()
```

Plot standardized individual intercepts and slopes for each clinical variable.
```{r plot_slopes_int}
mod %>% 
  dplyr::select(ID, ClinDx2, ends_with("slope"), ends_with("intercept")) %>% 
  gather(key=key, value=value, ECAS_Lang_slope:LMN_adj_intercept, -ID, -ClinDx2) %>% 
  separate(key, into = c("Test", "Stat"), sep="_(?=[^_]+$)") %>%
  ggplot(., aes(Test,value))+
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(color=Test), size=0.4, alpha=0.9) +
  facet_wrap(~Stat, scales = "free_y")+
  theme_minimal()+
  theme(legend.position = "none", axis.text.x = element_text(angle = 90))
```

Order the dataframe of intercepts and slopes according to clinical diagnosis, and then in descending order of ALSFRS-R intercept.
```{r cleaned_CReATe_data}
mod$ClinDx2 <- factor(mod$ClinDx2, levels = c("ALS", "ALSci", "ALS-FTD", "PLS", "PMA"))

cleaned_CReATe_data <- mod %>% 
  arrange(ClinDx2, desc(ALSFRS_TotalR_DM_intercept)) 

write.csv(cleaned_CReATe_data, "data_notshared/cleaned_CReATe_data.csv")
```

Create a heatmap to visualize intercepts and slopes across the CReATe PGB cohort (Figure 1A).
```{r figure_1A}
vars2 <- tidyselect::vars_select(names(cleaned_CReATe_data), ALSFRS_TotalR_DM_intercept, UMN_adj_intercept, LMN_adj_intercept, ends_with("intercept"), ALSFRS_TotalR_DM_slope, UMN_adj_slope, LMN_adj_slope, ends_with("slope"))

df <- cleaned_CReATe_data %>% 
  dplyr::select(all_of(vars2)) %>% 
  t()

rownames(df) = rep(c("ALSFRS-R Total", "UMN", "LMN", "Language", "Verbal Fluency", "Executive", "Memory", "Visuospatial", "ALS-Specific", "ALS-NonSpecific", "ECAS Total"),2)

#heatmap colors
col = colorRamp2(seq(min(df), max(df), length = 3), c("blue", "#EEEEEE", "red"), space = "RGB")

#column annotation
column_annot <- cleaned_CReATe_data %>% 
  dplyr::select(ClinDx2) %>% 
  rename(`Clinical Diagnosis` = ClinDx2) %>% 
  as.matrix(nrow=1) %>% 
  as.data.frame()

column_annot$`Clinical Diagnosis` <- factor(column_annot$`Clinical Diagnosis`, levels = c("ALS", "ALSci", "ALS-FTD", "PLS", "PMA"))

column_col = list("Clinical Diagnosis" = c("ALS" = "salmon", "ALSci" = "plum4",  "ALS-FTD" = "steelblue1", "PLS" = "gold3", "PMA" = "springgreen3"))

#row annotation
row_annot <- data.frame(c(rep(c("Adjusted Baseline"), 11), rep(c("Rate of Decline"), 11)))
names(row_annot) <- "Metric"

row_col = list(
  Metric = c("Adjusted Baseline" = "turquoise3", "Rate of Decline" = "orchid1"))
  
# Create the heatmap annotation
column_ha <- HeatmapAnnotation(df = column_annot, col = column_col)
row_ha <- rowAnnotation(df = row_annot, col = row_col)

# Combine the heatmap and the annotation
heatmap1 <- Heatmap(df, name = "Standard Deviation",
        top_annotation = column_ha, left_annotation = row_ha, cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = FALSE, row_names_gp = gpar(fontsize = 8), row_split = row_annot, row_title = " ", column_split = column_annot, column_title = " ", width = unit(4, "in"), height = unit(3, "in"))

heatmap1

png("Figure_1A.png", width = 8, height = 4.5, units = "in", res = 500)
heatmap1
dev.off()
```

Create a heatmap to visualize correlations between intercepts and slopes across clinical variables for the CReATe PGB cohort (Figure 1B).
```{r figure_1B}
df <- cleaned_CReATe_data %>% 
  dplyr::select(vars2) %>% 
  as.matrix()

colnames(df) = rep(c("ALSFRS-R Total", "UMN", "LMN", "Language", "Verbal Fluency", "Executive", "Memory", "Visuospatial", "ALS-Specific", "ALS-NonSpecific", "ECAS Total"),2)

df <- round(cor(df, method = c("spearman")),2)

# Get lower triangle of the correlation matrix
get_lower_tri <- function(df){
  df[upper.tri(df)]<- NA
  return(df)
}

lower_tri <- get_lower_tri(df)
  
# Create the heatmap annotation
column_ha <- HeatmapAnnotation(df = row_annot, col = row_col)
row_ha <- rowAnnotation(df = row_annot, col = row_col, show_legend = FALSE)

# Combine the heatmap and the annotation
heatmap2 <- Heatmap(lower_tri, name = "Spearman's R",
        top_annotation = column_ha, left_annotation = row_ha, cluster_rows = FALSE, cluster_columns = FALSE, row_split = row_annot, column_split = row_annot, row_title = " ", column_title = " ",row_names_gp = gpar(fontsize = 8), column_names_gp = gpar(fontsize = 8),width = unit(3, "in"), height = unit(3, "in") )

heatmap2

png("Figure_1B.png", width = 8, height = 4.5, units = "in", res = 500)
heatmap2
dev.off()
```

Create a heatmap to visualize individual values for genetic variables across the CReATe PGB cohort (Figure 1C).
```{r figure_1C}
df <- cleaned_CReATe_data %>% 
  dplyr::select(starts_with("rs"), Male, C9ORF72, Other_Mutation) %>%
  mutate_all(as.character) %>% 
  as.matrix() %>% 
  t()

#heatmap colors
colors = structure(c("dodgerblue3", "lightgoldenrod1", "firebrick3"), names = c("0", "1", "2"))

#annotation
column_ha <- HeatmapAnnotation(df = column_annot, col = column_col)

# Combine the heatmap and the annotation
heatmap3 <- Heatmap(df, name = "# Minor Alleles", col = colors,
        top_annotation = column_ha, cluster_rows = FALSE, cluster_columns = FALSE, show_column_names = FALSE, row_names_gp = gpar(fontsize = 6), column_split = column_annot, column_title = " ", width = unit(4, "in"), height = unit(3.5, "in"))

heatmap3

png("Figure_1C.png", width = 8, height = 4.5, units = "in", res = 500)
heatmap3
dev.off()
```

Combine three heatmaps and include in a single PDF (Figure 1).
```{r Figure1}
plots <- list(
  rasterGrob(readPNG("Figure_1A.png", native = FALSE),
    interpolate = FALSE), 
  rasterGrob(readPNG("Figure_1B.png", native = FALSE),
    interpolate = FALSE),
  rasterGrob(readPNG("Figure_1C.png", native = FALSE),
    interpolate = FALSE)
  )

pdf("Figure_1.pdf")
do.call(grid.arrange, c(plots, ncol = 1))
dev.off()
```

Plot principal components 1 and 2 from the PCA conducted in the CReATe PGB cohort (Appendix Figure S1).
```{r Appendix_Figure_S1}
pcplot <- ggplot(cleaned_CReATe_data, aes(PC1, PC2)) +
  geom_point() + 
  theme_minimal()

pdf("Appendix_Figure_S1.pdf")
pcplot
dev.off()
```


